A generalized definition of multidimensional item response theory parameters

### 2024-07-12 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds pending references

Multidimensional item response theory (MIRT) models have been widely used since
their inception.
The so-called *Multidimensional 2-Parameter Logistic*
[`r MODEL_ACRONYM`, @mckinley_extension_1983] model is among the most important
and widely used ones.
This is partly due to its close relationship to the common factor model
[@mcdonald_test_1999], but also to the formal tractability of its formulation.<!--# Ref. needed? -->
It consists of a generalization of the 2-parameter logistic model
[@birnbaum_latent_1968] to more than one latent trait
[@mckinley_extension_1983].
Its usefulness lies in describing the variability of the population when
the correct response does not depend on a single trait;
more specifically, when a deficit in a certain trait
and a surplus in another one cancel out their effects. Because of this property,
it is sometimes referred to as a *compensatory* model.

@@

The representation of a latent space where the traits are correlated is more
accurate on a geometrical coordinate system with non-orthogonal axes and
non-standard units, often referred to as
*general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the basis vectors and the axes they yield
have a meaningful interpretation (albeit counterintuitive, as we will see)
in a multivariate statistical sense.
The general case of MIRT model estimation implies estimating possibly
non-standard latent dimensions and non-null covariances among them
(a more realistic case than an identity matrix).
Assuming orthonormality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
[e.g., @thielmann_comparability_2022].
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
interpreting the latent dimensions as substantive constructs is not a goal.
As we will argue, the usual definitions of the multidimensional difficulty and
discrimination parameters do not generalize to the non-orthonormal geometry
necessary to accurately represent correlated latent spaces.

>>>>

Multidimensional item response theory (MIRT) models have been widely used since
their inception.
The so-called *Multidimensional 2-Parameter Logistic*
[`r MODEL_ACRONYM`, @mckinley_extension_1983] model is among the most important
and widely used ones.
This is partly due to its close relationship to the common factor model
[@mcdonald_test_1999], but also to the formal tractability of its formulation.
It consists of a generalization of the 2-parameter logistic model
[@birnbaum_latent_1968] to more than one latent trait
[@mckinley_extension_1983].
Its usefulness lies in describing the variability of the population when
the correct response does not depend on a single trait;
more specifically, when a deficit in a certain trait
and a surplus in another one cancel out their effects. Because of this property,
it is sometimes referred to as a *compensatory* model.

@@

The representation of a latent space where the traits are correlated is more
accurate on a geometrical coordinate system with non-orthogonal axes
[@ackerman_multidimensional_2005-1] and (possibly) non-standard units,
often referred to as *general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the basis vectors and the axes they yield
have a meaningful interpretation (albeit counterintuitive, as we will see)
in a multivariate statistical sense.
The general case of MIRT model estimation implies estimating possibly
non-standard latent dimensions and non-null covariances among them
(a more realistic case than an identity matrix).
Assuming orthonormality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
[e.g., @thielmann_comparability_2022].
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g., multigroup equating) even when
interpreting the latent dimensions as substantive constructs is not a goal
[for for multidimensional linking and equating procedures,
see @reckase_linking_2009].
As we will argue, the usual definitions of the multidimensional difficulty and
discrimination parameters do not generalize to the non-orthonormal geometry
necessary to accurately represent correlated latent spaces.
